---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# multidesign

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/bbuchsbaum/multidesign/graph/badge.svg)](https://app.codecov.io/gh/bbuchsbaum/multidesign)
<!-- badges: end -->

[Getting started](vignettes/Introduction.Rmd) ·
[Class overview](vignettes/class_overview.Rmd) ·
[Changelog](NEWS.md) ·
[Issue tracker](https://github.com/bbuchsbaum/multidesign/issues)

`multidesign` is an R package for multivariate experimental data where the matrix alone is not enough. It keeps observation-level design metadata and variable-level metadata attached to the data, so subsetting, grouping, splitting, and cross-validation stay aligned.

> **Status:** Pre-release and source-only. The API may change before the first
> stable release.

## Installation

`multidesign` is not on CRAN yet. Install the development version from GitHub:

```r
install.packages("remotes")
remotes::install_github("bbuchsbaum/multidesign")
```

## Quick start

```{r quick-start}
library(multidesign)

X <- matrix(seq_len(24), nrow = 6)
md <- multidesign(
  X,
  tibble::tibble(condition = rep(c("face", "house"), each = 3)),
  tibble::tibble(network = rep(c("visual", "control"), each = 2))
)

face_visual <- md |>
  subset(condition == "face") |>
  select_variables(network == "visual")

xdata(face_visual)
```

The matrix, row design, and column metadata stay synchronized through both
operations. This is the ordinary single-block workflow; use `hyperdesign` when
the data naturally contains several related blocks.

## Why use `multidesign`?

- Keep the data matrix, row design, optional column metadata, and explicit cell-observation mask together.
- Filter observations and variables without manually realigning metadata or masks.
- Split, summarize, and build cross-validation folds from experimental design variables.
- Represent corresponding entities across blocks without changing the package's row-centric matrix orientation.
- Scale to multi-subject, multi-session, or lazy-loading workflows with `hyperdesign` and `multiframe`.

## Core objects

| Class | Use it when... |
| --- | --- |
| `multidesign` | You have one data matrix plus row-wise design information and optional column metadata. |
| `hyperdesign` | You have multiple related blocks, optionally linked by corresponding entity IDs. |
| `multiframe` | Your observations are expensive to materialize and you want lazy evaluation. |
| `multiblock` | You need lower-level stacked matrix operations without the full design abstraction. |

Most workflows start with `multidesign` and move to `hyperdesign` when the data naturally breaks into subject-, session-, or modality-level blocks.

## Corresponding rows and partial observations

Rows usually represent trials, scans, or other independent observations. They
can also represent known entities—such as landmarks or tasks—that recur across
blocks. A `hyperdesign` can declare that second role with an entity key:

```{r correspondence}
shape_a <- multidesign(
  matrix(c(NA, 0, 1, 0, 0, 1), ncol = 2, byrow = TRUE),
  tibble::tibble(landmark = c("A", "B", "C")),
  cells = matrix(
    c(TRUE, TRUE, TRUE, FALSE, FALSE, FALSE),
    ncol = 2,
    byrow = TRUE
  )
)
shape_b <- multidesign(
  matrix(c(1, 0, 0, 1, 1, 1), ncol = 2, byrow = TRUE),
  tibble::tibble(landmark = c("B", "C", "D"))
)

shapes <- hyperdesign(
  list(shape_a = shape_a, shape_b = shape_b),
  id = "landmark",
  space = "common"
)

corr <- correspondence(shapes)
corr$global_ids
do.call(rbind, corr$row_map)
corr$overlap$pair_n

aligned <- align_by_id(shapes)
aligned$observed
aligned$cells[, , "shape_a"]
aligned$x[, , "shape_a"]

correspondence(shapes, sort_ids = TRUE)$global_ids

stopifnot(all(vapply(seq_along(shapes), function(i) {
  identical(corr$global_ids[corr$row_map[[i]]], corr$ids[[i]])
}, logical(1))))
```

The aligned result distinguishes three states that can all display as `NA` in
`aligned$x`: landmark D is absent from `shape_a` (`observed` is `FALSE`),
landmark C is present but has an all-`FALSE` cell-mask row, and landmark A's
first coordinate is observed even though its stored value is `NA`. The `fill`
argument changes only the displayed value for absent or masked cells; consult
`observed` and `cells` for their meaning. A missing value in `x` never creates
a mask implicitly.

This is a role of the row key, not a second matrix orientation. **Do not store
features × observations in `x` to satisfy a left-action formula.** A solver that
uses left-action notation should transpose at its boundary.

The declarations answer different questions: `common_vars` records design
columns shared by every block (schema), `id` names values that join rows
(correspondence), and `space` says whether columns are shared axes (`"common"`)
or block-specific variables (`"block"`). `as_multidesign()` always row-stacks;
it never joins by `id`, and it refuses `space = "block"`. `align_by_id()` is
the explicit join: it returns common-space arrays on the global entity
universe, but it does not fit an alignment model.

Cell masks move with ordinary row and column subsets. When masked rows or
duplicate entity keys must be combined, aggregation is explicit:
`summarize_by(x, group, aggregate = "mean")` and
`hyperdesign(blocks, id = "landmark", aggregate = "mean")`. A custom
scalar-returning function may replace `"mean"`. Stacking promotes an unmasked
block to all observed when another block has a mask. Preprocessing and
dimensionality reduction reject partial masks because those operations do not
declare how the mask itself transforms.

These contracts let downstream packages validate data for generalized
Procrustes analysis without putting transforms, gauges, fitting, or
certification in `multidesign`. Likewise, `fold_over(shapes, landmark)` holds an
entity out from all blocks where it occurs; it is an ordinary data split, not a
specialized Bai–Bartoli/LBW cross-validation engine. Use `positional = TRUE`
only when equal row positions really are the correspondence and that fragile
assumption is intentional.

## Design-aware cross-validation

```r
set.seed(42)

# Simulate a trial-by-feature matrix
X <- matrix(rnorm(40 * 8), nrow = 40, ncol = 8)

# Design metadata for each observation
design_df <- tibble::tibble(
  condition = rep(c("face", "house"), each = 20),
  run = rep(1:4, each = 10),
  trial = 1:40
)

# Metadata for each variable / column
column_df <- tibble::tibble(
  roi = paste0("ROI_", 1:8),
  network = rep(c("visual", "frontal"), each = 4)
)

# Build a multidesign object
md <- multidesign(X, design_df, column_df)

# Row-wise and column-wise access stay synchronized
face_trials <- subset(md, condition == "face")
visual_rois <- select_variables(md, network == "visual")

# Summarize by an experimental factor
condition_means <- summarize_by(md, condition)

# Create design-aware cross-validation folds
folds <- fold_over(md, run)

# Run a simple CV pipeline
cv <- cross_validate(
  folds,
  fit_fn = function(analysis) {
    list(mean = colMeans(xdata(analysis)))
  },
  score_fn = function(model, assessment) {
    mat <- xdata(assessment)
    pred <- matrix(
      model$mean,
      nrow = nrow(mat),
      ncol = ncol(mat),
      byrow = TRUE
    )
    c(mse = mean((mat - pred)^2))
  }
)

summary(cv)
```

## Typical workflow

1. Create a `multidesign` from a matrix and a design table.
2. Subset rows by design variables and columns by variable metadata.
3. Summarize or split by condition, run, subject, or any other design factor.
4. Generate folds with `fold_over()` or `cv_rows()`.
5. Pass those folds into `cross_validate()` with your model-fitting and scoring functions.
6. Move to `hyperdesign` when the analysis spans multiple related blocks.

## Learn more

- [Package website](https://bbuchsbaum.github.io/multidesign/)
- [Getting started vignette](vignettes/Introduction.Rmd)
- [Class overview vignette](vignettes/class_overview.Rmd)
- [Changelog](NEWS.md)
- [Issue tracker](https://github.com/bbuchsbaum/multidesign/issues)

## Albers theme

This package uses the albersdown theme. Existing vignette theme hooks are replaced so `albers.css` and local `albers.js` render consistently on CRAN and GitHub Pages. The defaults are configured via `params$family` and `params$preset` (family = "ochre", preset = "homage"). The pkgdown site uses `template: { package: albersdown }`.
