---
title: Class Reference and Design Patterns
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
vignette: >
  %\VignetteIndexEntry{Class Reference and Design Patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: ochre
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

---

```{r setup, include=FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(123)
library(multidesign)
library(tibble)
library(dplyr)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

# Overview

The `multidesign` package provides a family of S3 classes for representing multivariate experimental data. This vignette serves as a reference guide, documenting each class, its purpose, key methods, and typical usage patterns.

## Class hierarchy

```
multidesign ─────────────────────── Core class: matrix + row design + column design
     │
     ├── reduced_multidesign ────── After dimensionality reduction
     │
     └── hyperdesign ────────────── Collection of multidesign blocks
              │
              └── as_multidesign() → Collapse back to single multidesign

multiframe ──────────────────────── Lazy-evaluated observations + design

multiblock ──────────────────────── Stacked matrices (row or column oriented)

foldlist ────────────────────────── Cross-validation fold container

cv_result ───────────────────────── Cross-validation results
```

## Choosing the right class

```
Do you have multiple subjects/sessions/modalities?
├── YES → hyperdesign
└── NO → Is your data expensive to load?
         ├── YES → multiframe
         └── NO → multidesign

Do you need to stack matrices without design info?
└── YES → multiblock
```

# multidesign

## Purpose

A `multidesign` is the fundamental class in the package. It binds together:

- **Data matrix** (`x`): rows are observations, columns are variables
- **Row design** (`design`): a tibble describing each observation
- **Column design** (`column_design`): a tibble describing each variable (optional)

## When to use

- Standard single-block multivariate analysis
- You need to subset or split data by experimental factors
- You want cross-validation that respects design structure
- Your variables have metadata (ROI names, feature types, etc.)

## When NOT to use
- You have multiple subjects/sessions → use `hyperdesign`
- Your data is too large to load at once → use `multiframe`
- You just need to stack matrices without design info → use `multiblock`

## Construction

```{r}
# Data matrix: 30 observations x 10 variables
X <- matrix(rnorm(30 * 10), 30, 10)

# Row design: what describes each observation
row_design <- tibble(
  condition = rep(c("A", "B", "C"), each = 10),
  block = rep(1:3, 10)
)

# Column design: what describes each variable (optional)
col_design <- tibble(
  variable = paste0("V", 1:10),
  group = rep(c("group1", "group2"), 5)
)

# Create multidesign
md <- multidesign(X, row_design, col_design)
md
```

## Key methods

| Method | Description |
|--------|-------------|
| `xdata(x)` | Extract the data matrix |
| `design(x)` | Extract row design (without internal `.index`) |
| `column_design(x)` | Extract column design |
| `subset(x, expr)` | Filter rows by design expression |
| `split(x, ...)` | Split into list by design variables |
| `select_variables(x, ...)` | Filter columns by column_design expression |
| `summarize_by(x, ...)` | Compute group summaries |
| `fold_over(x, ...)` | Create cross-validation folds |
| `cv_rows(x, rows)` | Create folds from explicit row indices |
| `split_indices(x, ...)` | Get row indices grouped by design variables |

## Examples

```{r}
# Extract components
dim(xdata(md))
head(design(md), 3)
column_design(md)

# Subset by design
md_A <- subset(md, condition == "A")
nrow(xdata(md_A))

# Select variables by column design
md_g1 <- select_variables(md, group == "group1")
ncol(xdata(md_g1))

# Summarize by condition
md_means <- summarize_by(md, condition)
design(md_means)
dim(xdata(md_means))

# Create CV folds by condition
folds <- fold_over(md, condition)
length(folds)

# Preserve original source row ids when needed downstream
folds_with_ids <- fold_over(md, condition, preserve_row_ids = TRUE)
head(folds_with_ids[[1]]$assessment$design$.orig_index)
```

---

# reduced_multidesign
## Purpose

A `reduced_multidesign` extends `multidesign` to store the result of dimensionality reduction (e.g., PCA). It contains the reduced data and the projector used for the transformation.

## When to use

- After applying `reduce()` to a multidesign
- When you need to project new data using the same transformation

## Construction

```{r}
# Note: requires multivarious package for full functionality
# Basic structure shown here
md_small <- multidesign(
  matrix(rnorm(50 * 20), 50, 20),
  tibble(group = rep(letters[1:5], 10))
)

# reduce() performs PCA by default
# rd <- reduce(md_small, nc = 3)
# rd
```

The `reduce()` function requires the `multivarious` package. When called, it returns a `reduced_multidesign` with:

- Reduced data matrix (observations × components)
- Original design information
- The projector object for transforming new data

---

# hyperdesign

## Purpose

A `hyperdesign` manages multiple `multidesign` objects (called "blocks") that share common design variables. This is the natural structure for multi-subject, multi-session, or multi-modal experiments.

## When to use

- Multiple subjects, each with their own data matrix
- Multiple sessions or runs that should be analyzed together
- Multi-view data (e.g., fMRI + EEG from the same experiment)
- You want leave-one-block-out cross-validation

## When NOT to use

- Single-subject analysis → use `multidesign`
- Blocks don't share any design structure → consider separate `multidesign` objects

## Construction

```{r}
# Create individual subject data
subj1 <- multidesign(
  matrix(rnorm(20 * 5), 20, 5),
  tibble(condition = rep(c("X", "Y"), 10), trial = 1:20)
)

subj2 <- multidesign(
  matrix(rnorm(20 * 5), 20, 5),
  tibble(condition = rep(c("X", "Y"), 10), trial = 1:20)
)

subj3 <- multidesign(
  matrix(rnorm(20 * 5), 20, 5),
  tibble(condition = rep(c("X", "Y"), 10), trial = 1:20)
)

# Combine into hyperdesign
hd <- hyperdesign(
  list(subj1, subj2, subj3),
  block_names = c("subject_1", "subject_2", "subject_3")
)
hd
```

## Key methods

| Method | Description |
|--------|-------------|
| `xdata(x)` / `xdata(x, block=i)` | Extract data (all blocks or specific block) |
| `design(x)` / `design(x, block=i)` | Extract design (all blocks or specific block) |
| `column_design(x)` | Extract column design |
| `subset(x, expr)` | Filter all blocks by expression |
| `select_variables(x, ...)` | Filter columns in all blocks |
| `fold_over(x)` | Leave-one-block-out folds |
| `fold_over(x, var)` | Within-block folds by variable |
| `cv_rows(x, rows)` | Create synchronized row-index folds |
| `as_multidesign(x)` | Collapse to single multidesign |
| `block_indices(x, i)` | Get indices for block i |

## Examples

```{r}
# Access individual blocks
length(hd)
xdata(hd, block = 1)[1:3, ]

# Access all designs
designs <- design(hd)
length(designs)

# Leave-one-block-out cross-validation
loso <- fold_over(hd)
length(loso)

# Each fold: analysis is a hyperdesign (n-1 blocks), assessment is a multidesign (1 block)
f1 <- loso[[1]]
length(f1$analysis)
class(f1$assessment)

# Within-block CV by condition
within_folds <- fold_over(hd, condition)
length(within_folds)

# Explicit synchronized row holdouts
row_folds <- cv_rows(
  hd,
  rows = list(list(subject_1 = 1:2, subject_2 = 1:2)),
  preserve_row_ids = TRUE
)
row_folds[[1]]$assessment[[1]]$design$.orig_index

# Collapse to single multidesign
md_collapsed <- as_multidesign(hd, .id = "subject")
nrow(xdata(md_collapsed))
table(design(md_collapsed)$subject)
```

## Alternative construction: from data frame

```{r}
# If your data is in a wide-format data frame
wide_df <- tibble(
  subject = rep(1:2, each = 5),
  condition = rep(c("A", "B", "A", "B", "A"), 2),
  v1 = rnorm(10),
  v2 = rnorm(10),
  v3 = rnorm(10)
)

hd_from_df <- df_to_hyperdesign(
  wide_df,
  design_vars = c("condition"),
  x_vars = c("v1", "v2", "v3"),
  split_var = "subject"
)
hd_from_df
```

---

# multiframe

## Purpose

A `multiframe` is similar to a `multidesign` but uses lazy evaluation for observations. Each observation is stored as a function that returns the data when called. This is useful when data is expensive to load or too large to hold entirely in memory.

## When to use

- Data files are large and you only need subsets at a time
- Observations are loaded from disk on demand
- You want to manipulate design metadata without loading all data

## When NOT to use

- Data fits easily in memory → use `multidesign` (simpler)
- You need to operate on the full matrix frequently → use `multidesign`

## Construction

```{r}
# From a matrix (each row becomes a lazy observation)
X <- matrix(1:40, 10, 4)
design_info <- tibble(
  group = rep(c("A", "B"), 5),
  id = 1:10
)

mf <- multiframe(X, design_info)
mf

# From a list (each element is a lazy observation)
obs_list <- list(
  matrix(1:12, 3, 4),
  matrix(13:24, 3, 4),
  matrix(25:36, 3, 4)
)
mf_list <- multiframe(obs_list, tibble(condition = c("X", "Y", "Z")))
```

## Key methods

| Method | Description |
|--------|-------------|
| `design(x)` | Extract design (without `.obs` and `.index`) |
| `xdata(x)` | Materialize all observations into a matrix |
| `subset(x, expr)` | Filter by design (lazy observations preserved) |
| `split(x, ...)` | Split by design variables |
| `split_indices(x, ...)` | Get indices grouped by design |
| `summarize_by(x, ...)` | Compute group summaries |
| `fold_over(x, ...)` | Create cross-validation folds |
| `cv_rows(x, rows)` | Create folds from explicit row indices |

## Examples

```{r}
# Design without internal columns
design(mf)

# Materialize all data (evaluates all lazy observations)
mat <- xdata(mf)
dim(mat)

# Subset (lazy observations remain lazy)
mf_A <- subset(mf, group == "A")
nrow(mf_A$design)

# Access individual observations lazily
mf$design$.obs[[1]]()

# Create CV folds
mf_folds <- fold_over(mf, group)
length(mf_folds)

# Preserve original source row ids in fold outputs
mf_folds_with_ids <- fold_over(mf, group, preserve_row_ids = TRUE)
mf_folds_with_ids[[1]]$assessment$design$.orig_index
```

---

# multiblock

## Purpose

A `multiblock` represents a collection of matrices that share either a row dimension (column-stacked) or a column dimension (row-stacked). It's a lower-level structure than `hyperdesign`, useful when you need to work with the combined matrix directly.

## When to use

- Matrices that should be stacked into one large matrix
- Block-wise operations on combined data
- Input to multi-block statistical methods

## When NOT to use

- You need design metadata → use `hyperdesign`
- Matrices don't share a dimension → can't be stacked

## Construction

```{r}
# Column-stacked: same number of rows, different columns
A1 <- matrix(1:12, 4, 3)
A2 <- matrix(13:20, 4, 2)
mb_col <- multiblock(list(A1, A2))
mb_col

# Row-stacked: same number of columns, different rows
B1 <- matrix(1:10, 2, 5)
B2 <- matrix(11:25, 3, 5)
mb_row <- multiblock(list(B1, B2))
mb_row
```

## Key methods

| Method | Description |
|--------|-------------|
| `is_cstacked(x)` | Check if column-stacked |
| `is_rstacked(x)` | Check if row-stacked |
| `block_indices(x, i)` | Get indices for block i |
| `t(x)` | Transpose (switches orientation) |

## Examples

```{r}
# Check orientation
is_cstacked(mb_col)
is_rstacked(mb_row)

# Get block indices
block_indices(mb_col, 1)
block_indices(mb_col, 2)

# Transpose switches orientation
mb_t <- t(mb_col)
is_rstacked(mb_t)
```

---

# foldlist

## Purpose

A `foldlist` is returned by `fold_over()` or `cv_rows()` and contains cross-validation folds. Each fold has an analysis set (training data) and an assessment set (test data).
## Structure

Each element of a foldlist contains:

- `analysis`: The training data (same class as input)
- `assessment`: The test data (same class as input, or `multidesign` for hyperdesign assessment)
- `held_out`: Information about what was held out (optional)

When folds are created with `preserve_row_ids = TRUE`, the fold outputs also
carry a reserved `.orig_index` design column, and `held_out$row_ids` records
the same source-row identities explicitly.

## Examples

```{r}
md <- multidesign(
  matrix(rnorm(40), 10, 4),
  tibble(group = rep(c("A", "B"), 5))
)

folds <- fold_over(md, group)
folds

# Access a specific fold
f1 <- folds[[1]]
names(f1)

# Training and test sets
nrow(xdata(f1$analysis))
nrow(xdata(f1$assessment))

# The foldframe attribute contains fold metadata
attr(folds, "foldframe")

# Explicit row folds work too
row_folds <- cv_rows(md, rows = list(1:2), preserve_row_ids = TRUE)
row_folds[[1]]$held_out$row_ids
```

---

# cv_result

## Purpose

A `cv_result` object stores the results of cross-validation executed via `cross_validate()`. It contains per-fold scores and provides methods for summarizing performance.

## Structure

- `scores`: A tibble with a `.fold` column and one or more metric columns
- `call`: The original call to `cross_validate()` (optional)

## Key methods

| Method | Description |
|--------|-------------|
| `print(x)` | Display fold count and metric summaries |
| `summary(x)` | Return tibble with mean/sd/min/max/median per metric |

## Examples

```{r}
# Run cross-validation
md <- multidesign(
  matrix(rnorm(60), 15, 4),
  tibble(condition = rep(c("A", "B", "C"), 5))
)
folds <- fold_over(md, condition)

cv_res <- cross_validate(
  folds,
  fit_fn = function(analysis) {
    list(center = colMeans(xdata(analysis)))
  },
  score_fn = function(model, assessment) {
    mat <- xdata(assessment)
    pred <- matrix(model$center, nrow = nrow(mat), ncol = ncol(mat), byrow = TRUE)
    c(mse = mean((mat - pred)^2), mae = mean(abs(mat - pred)))
  }
)

# Print summary
cv_res

# Detailed summary
summary(cv_res)

# Access raw scores
cv_res$scores
```

---

# Combining and converting between classes

## bind_multidesign: Combine multiple multidesign objects

```{r}
md1 <- multidesign(matrix(1:10, 5, 2), tibble(g = rep("A", 5)))
md2 <- multidesign(matrix(11:20, 5, 2), tibble(g = rep("B", 5)))

# Bind with source tracking
combined <- bind_multidesign(md1, md2, .id = "source")
design(combined)

# Also accepts hyperdesign (expands blocks first)
hd <- hyperdesign(list(md1, md2))
md3 <- multidesign(matrix(21:30, 5, 2), tibble(g = rep("C", 5)))
combined2 <- bind_multidesign(hd, md3)
nrow(xdata(combined2))
```

## as_multidesign: Collapse hyperdesign

```{r}
hd <- hyperdesign(
  list(
    multidesign(matrix(1:10, 5, 2), tibble(cond = c("X", "X", "Y", "Y", "Y"))),
    multidesign(matrix(11:20, 5, 2), tibble(cond = c("X", "X", "Y", "Y", "Y")))
  ),
  block_names = c("block1", "block2")
)

# Collapse to single multidesign
md_flat <- as_multidesign(hd, .id = "block")
design(md_flat)
```

---

# Method availability summary

| Method | multidesign | hyperdesign | multiframe | multiblock |
|--------|:-----------:|:-----------:|:----------:|:----------:|
| `xdata()` | ✓ | ✓ | ✓ | - |
| `design()` | ✓ | ✓ | ✓ | - |
| `column_design()` | ✓ | ✓ | - | - |
| `subset()` | ✓ | ✓ | ✓ | - |
| `split()` | ✓ | - | ✓ | - |
| `select_variables()` | ✓ | ✓ | - | - |
| `summarize_by()` | ✓ | - | ✓ | - |
| `fold_over()` | ✓ | ✓ | ✓ | - |
| `cv_rows()` | ✓ | ✓ | ✓ | - |
| `split_indices()` | ✓ | - | ✓ | - |
| `block_indices()` | - | ✓ | - | ✓ |
| `is_cstacked()` | - | - | - | ✓ |
| `is_rstacked()` | - | - | - | ✓ |

---

# Session info

```{r}
sessionInfo()
```
