---
title: "Class Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Class Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

This vignette provides a high level overview of the S3 classes implemented in
**multidesign**.  Each class encapsulates a different way of representing
multivariate data together with its associated experimental design.  The goal is
to make it easy to switch between single data matrices and more complex
multi–block structures while keeping a consistent interface.

# `multidesign`

A `multidesign` object links a multivariate matrix `x` with a data frame of
design variables.  Every row of `x` corresponds to a row in the design.  This is
useful for standard single–block analyses where the observations share a common
set of descriptors.

```{r}
X <- matrix(rnorm(20 * 5), 20, 5)
Y <- tibble(condition = rep(letters[1:4], 5))
md <- multidesign(X, Y)
md
```

# `reduced_multidesign`

The `reduced_multidesign` class extends `multidesign` and stores the result of a
projection or dimension–reduction step.  It contains the reduced data and the
projector used to obtain it.

```{r}
rd <- reduce(md, nc = 2)
rd
```

# `hyperdesign`

A `hyperdesign` is a collection of `multidesign` objects that share one or more
common design variables.  It is suited for multiblock or multiview settings
where each block represents a separate matrix but the blocks are related through
shared variables.

```{r}
hd <- hyperdesign(list(md, md))
hd
```

# `multiblock`

The `multiblock` class represents a single matrix composed of stacked sub
matrices that share either a row or column dimension.  It is a lower level
format than `hyperdesign` and is mainly used when the blocks form one large
matrix.

```{r}
mb <- multiblock(list(X, X))
mb
```

# `multiframe`

`multiframe` stores the design as its main component and represents each
observation as a delayed computation via the `observation` class.  This is
helpful when the underlying data are expensive to load into memory but the
metadata can be manipulated using dplyr verbs.

```{r}
mf <- multiframe(list(X[1:5,], X[6:10,]), tibble(run = c(1,2)))
mf
```

# `observation` and `observation_set`

`observation` objects are functions that return a single data slice when called.
An `observation_set` is simply a list of `observation` objects.  They provide a
light–weight way of lazily accessing individual observations from a larger data
source.

# `foldlist`

A `foldlist` is returned by `fold_over()` and represents a set of cross
validation folds.  Each element contains an analysis set and an assessment set,
both expressed using the same classes described above.

# Summary

These classes work together to support flexible representations of complex
experimental designs.  Starting with a simple `multidesign`, you can build
`hyperdesign` objects, perform cross–validation with `foldlist`, or defer data
loading with `multiframe` and `observation`.
