---
title: "Introduction to multidesign"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to multidesign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(multidesign)
library(tibble)
library(dplyr)
```

# Introduction

The `multidesign` package provides a flexible framework for working with multiple related datasets, particularly in the context of multiblock and multiview data analysis. The package introduces a set of data structures, each serving a specific purpose in organizing and managing complex experimental data.

# Core Data Structures

## 1. Observation Sets (`observation_set`)

The most basic building block is the `observation_set`, which provides a unified interface for accessing individual observations:

```{r}
# Create from a matrix (each row is an observation)
X_mat <- matrix(rnorm(20), 4, 5)
obs_mat <- obs_group(X_mat)

# Create from a list (each element is an observation)
X_list <- list(
  matrix(1:12, 3, 4),
  matrix(13:24, 3, 4),
  matrix(25:36, 3, 4)
)
obs_list <- obs_group(X_list)

# Access observations
obs_mat[[1]]  # First observation from matrix
obs_list[[1]] # First observation from list
```

Key features:
- Lazy evaluation of observations
- Unified access pattern regardless of underlying data structure
- Support for matrices, lists, and vectors

## 2. Multiframes (`multiframe`)

A `multiframe` combines observations with experimental design metadata:

```{r}
# Create observations
X <- list(
  matrix(rnorm(20), 4, 5),    # 4x5 matrix
  matrix(rnorm(20), 4, 5),    # 4x5 matrix
  matrix(rnorm(20), 4, 5)     # 4x5 matrix
)

# Create design information
Y <- tibble(
  condition = c("control", "treatment", "control"),
  subject = 1:3,
  block = c(1, 1, 2)
)

# Create multiframe
mf <- multiframe(X, Y)

# Split by condition
split_by_cond <- split(mf, condition)

# Compute means by condition
means_by_cond <- summarize_by(mf, condition)
```

Key features:
- Combines observations with design variables
- Supports splitting and summarizing by design variables
- Automatic dimension validation
- Lazy evaluation of observations

## 3. Multidesigns (`multidesign`)

A `multidesign` extends `multiframe` by adding support for column metadata and more sophisticated operations:

```{r}
# Create data matrix
X <- matrix(rnorm(100*20), 100, 20)

# Create design information
Y <- tibble(
  condition = rep(c("control", "treatment"), each=50),
  subject = rep(1:10, each=10)
)

# Create column metadata
col_info <- tibble(
  roi = paste0("region_", 1:20),
  hemisphere = rep(c("left", "right"), 10)
)

# Create multidesign
mds <- multidesign(X, Y, col_info)


```

Key features:
- Column metadata support
- Subset and split operations
- Cross-validation support via folding

## 4. Hyperdesigns (`hyperdesign`)

A `hyperdesign` manages multiple related `multidesign` objects:

```{r}
# Create multiple multidesign objects (e.g., for different subjects)
d1 <- multidesign(
  matrix(rnorm(10*20), 10, 20),
  tibble(run=1:10, condition=rep(c("A","B"), 5))
)

d2 <- multidesign(
  matrix(rnorm(10*20), 10, 20),
  tibble(run=1:10, condition=rep(c("A","B"), 5))
)

# Combine into hyperdesign
hd <- hyperdesign(list(d1, d2), block_names=c("subject1", "subject2"))

# Create cross-validation folds
folds <- fold_over(hd)
```

Key features:
- Manages multiple related datasets
- Preserves common design variables
- Cross-validation across blocks
- Block-wise operations

