---
title: Getting Started with multidesign
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
vignette: |
  %\VignetteIndexEntry{Getting Started with multidesign} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
params:
  family: ochre
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

---

```{r, include = FALSE}
knitr::opts_chunk$set(
 collapse = TRUE,
 comment = "#>"
)
set.seed(42)
```

```{r setup}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(multidesign)
library(tibble)
library(dplyr)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


# Why multidesign?

In experimental research—particularly neuroimaging, psychophysics, and behavioral studies—data often has a specific structure that standard R data frames don't capture well:

- **Observations** (trials, time points, scans) arranged as rows in a matrix
- **Variables** (brain regions, sensors, features) arranged as columns
- **Design metadata** (condition, subject, run, block) describing each observation
- **Variable metadata** (ROI name, hemisphere, sensor type) describing each column

Without a unified structure, you end up with separate objects for data and metadata that can easily get out of sync, manual bookkeeping for cross-validation splits, and difficulty handling multi-subject or multi-session experiments.

The `multidesign` package solves these problems by providing data structures that:

1. **Keep data and metadata together** — No more matching row indices between separate objects
2. **Support design-aware operations** — Split, subset, and summarize by experimental factors
3. **Handle multi-block data naturally** — Multiple subjects, sessions, or modalities
4. **Provide built-in cross-validation** — Proper stratification respecting your experimental design

# Quick reference: When to use each class

| Class | Use when... |
|-------|-------------|
| `multidesign` | You have a single data matrix with row-wise design information and optional column metadata |
| `hyperdesign` | You have multiple related matrices (e.g., different subjects or sessions) that share design structure |
| `multiframe` | Your data is expensive to load and you want lazy evaluation (load on demand) |
| `multiblock` | You have stacked matrices sharing a row or column dimension (low-level operations) |

Most users will start with `multidesign` and move to `hyperdesign` when working with multi-subject data.

# A complete workflow

Let's walk through a realistic analysis workflow using simulated neuroimaging-style data.

## Step 1: Create your data

Imagine a simple experiment: 2 conditions (face vs. house), 10 subjects, 20 trials per condition per subject, measuring activity in 50 brain regions.

```{r}
# Simulate data for one subject
n_trials <- 40
n_regions <- 50

# Create the data matrix (trials x brain regions)
X <- matrix(rnorm(n_trials * n_regions), n_trials, n_regions)

# Create the design information
# Each run has 10 trials: 5 face + 5 house
design_df <- tibble(
 condition = rep(rep(c("face", "house"), each = 5), 4),
 trial = 1:40,
 run = rep(1:4, each = 10)
)

# Create column metadata describing the brain regions
column_info <- tibble(
 region = paste0("ROI_", 1:n_regions),
 hemisphere = rep(c("left", "right"), 25),
 network = rep(c("visual", "frontal", "parietal", "temporal", "default"), 10)
)
```

## Step 2: Build a multidesign

Combine your data matrix, design information, and column metadata into a single object:
```{r}
mds <- multidesign(X, design_df, column_info)
mds
```

The `multidesign` object keeps everything synchronized. You can extract components with accessor functions:

```{r}
# Get the data matrix
head(xdata(mds)[, 1:5])

# Get the design (without internal .index column)
head(design(mds))

# Get column metadata
head(column_design(mds))
```

## Step 3: Explore and subset your data

Filter observations by design variables:

```{r}
# Keep only face trials
mds_face <- subset(mds, condition == "face")
mds_face

# Keep only run 1
mds_run1 <- subset(mds, run == 1)
nrow(xdata(mds_run1))
```

Select variables (columns) by their metadata:

```{r}
# Keep only visual network regions
mds_visual <- select_variables(mds, network == "visual")
ncol(xdata(mds_visual))
column_design(mds_visual)

# Keep only left hemisphere visual regions
mds_left_visual <- select_variables(mds, network == "visual" & hemisphere == "left")
ncol(xdata(mds_left_visual))
```

## Step 4: Summarize by design factors

Compute summary statistics grouped by design variables:

```{r}
# Mean activity per condition
mds_means <- summarize_by(mds, condition)
mds_means

# The result is a new multidesign with one row per group
dim(xdata(mds_means))
design(mds_means)
```

## Step 5: Split for detailed analysis

Split into separate multidesign objects:

```{r}
# Split by condition
by_condition <- split(mds, condition)
length(by_condition)
names(by_condition) <- c("face", "house")

# Each element is a complete multidesign
by_condition$face
```

## Step 6: Cross-validation

Create cross-validation folds that respect your experimental design:

```{r}
# Leave-one-run-out cross-validation
folds <- fold_over(mds, run)
folds

# Each fold contains analysis (training) and assessment (test) sets
fold1 <- folds[[1]]
nrow(xdata(fold1$analysis))
nrow(xdata(fold1$assessment))
```

If downstream code needs stable source-row identities, ask fold construction to
carry them through:

```{r}
folds_with_ids <- fold_over(mds, run, preserve_row_ids = TRUE)
folds_with_ids[[1]]$assessment$design$.orig_index
```

## Step 7: Run a cross-validation pipeline

Use `cross_validate()` to execute a complete CV workflow:

```{r}
# Define a simple "model": compute condition means on training data
fit_fn <- function(analysis) {
 means_by_cond <- summarize_by(analysis, condition)
 list(
   face_mean = xdata(means_by_cond)[1, ],
   house_mean = xdata(means_by_cond)[2, ]
 )
}

# Define scoring: correlation between predicted and actual means
score_fn <- function(model, assessment) {
 test_means <- summarize_by(assessment, condition)

 # Compare model predictions to test data
 face_cor <- cor(model$face_mean, xdata(test_means)[1, ])
 house_cor <- cor(model$house_mean, xdata(test_means)[2, ])

 c(face_r = face_cor, house_r = house_cor, mean_r = (face_cor + house_cor) / 2)
}

# Run cross-validation
cv_result <- cross_validate(folds, fit_fn, score_fn)
cv_result
```

Examine detailed results:

```{r}
# Get summary statistics
summary(cv_result)

# Access raw scores
cv_result$scores
```

# Working with multiple subjects: hyperdesign

Real experiments typically have multiple subjects. A `hyperdesign` manages multiple `multidesign` objects that share common design structure.

```{r}
# Create data for 3 subjects
subjects <- lapply(1:3, function(subj) {
 X_subj <- matrix(rnorm(n_trials * n_regions), n_trials, n_regions)

 # Add a subject-specific signal
 if (subj == 1) X_subj[1:20, 1:10] <- X_subj[1:20, 1:10] + 0.5
 if (subj == 2) X_subj[1:20, 1:10] <- X_subj[1:20, 1:10] + 0.3
 if (subj == 3) X_subj[1:20, 1:10] <- X_subj[1:20, 1:10] + 0.7

 multidesign(X_subj, design_df, column_info)
})

# Combine into a hyperdesign
hd <- hyperdesign(subjects, block_names = c("subj1", "subj2", "subj3"))
hd
```

## Leave-one-subject-out cross-validation

```{r}
# Without specifying variables, fold_over creates leave-one-block-out folds
loso_folds <- fold_over(hd)
length(loso_folds)

# Each fold: train on 2 subjects, test on 1
fold1 <- loso_folds[[1]]
length(fold1$analysis)
fold1$assessment
```

## Within-subject cross-validation

```{r}
# Fold by run within each subject
run_folds <- fold_over(hd, run)
length(run_folds)
```

## Explicit row-based cross-validation

When you already know the held-out rows, use `cv_rows()` to build a
`foldlist` directly:

```{r}
row_folds <- cv_rows(
  hd,
  rows = list(
    list(subj1 = 1:2, subj2 = 1:2),
    list(subj1 = 3:4, subj2 = 3:4)
  )
)

row_folds[[1]]$assessment
```

Use `preserve_row_ids = TRUE` here as well if external correspondence tables or
annotations are keyed by original rows:

```{r}
row_folds_with_ids <- cv_rows(
  hd,
  rows = list(list(subj1 = 1:2, subj2 = 1:2)),
  preserve_row_ids = TRUE
)

row_folds_with_ids[[1]]$assessment[[1]]$design$.orig_index
row_folds_with_ids[[1]]$held_out$row_ids
```

## Collapse to a single multidesign

If you need to analyze all subjects together:

```{r}
# Collapse hyperdesign to one multidesign
mds_all <- as_multidesign(hd, .id = "subject")
mds_all

# The .id column tracks which block each observation came from
table(design(mds_all)$subject)
```

# Next steps

- See `vignette("class_overview")
` for detailed documentation of each class
- Explore the function reference for methods like `bind_multidesign()`, `df_to_hyperdesign()`, and `reduce()`
- Check out the package website for additional examples

# Session info

```{r}
sessionInfo()
```
